---
title: "Warned vs Unwarned Symptoms"
subtitle: "Nocebo side effects — Saunders et al. (2025)"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    self-contained: true
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
---
```{r setup}
library(tidyverse)
library(readxl)
library(plotly)
library(here)

datapath <- here("data", "1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

df_long <- df %>%
  select(gid_pid, group, intervention_bin,
         starts_with("bgase_"), starts_with("agase_")) %>%
  pivot_longer(
    cols          = starts_with("bgase_") | starts_with("agase_"),
    names_to      = c("timing", "item"),
    names_pattern = "(bgase|agase)_(\\d+)"
  ) %>%
  pivot_wider(names_from = timing, values_from = value) %>%
  mutate(
    diff    = agase - bgase,
    item    = as.integer(item),
    warned  = item <= 2,
    symptom = case_when(
      item == 1  ~ "Headache",
      item == 2  ~ "Dizziness",
      item == 3  ~ "Nausea",
      item == 4  ~ "Stomach discomfort",
      item == 5  ~ "Breathing problems",
      item == 6  ~ "Palpitation",
      item == 7  ~ "Itching",
      item == 8  ~ "Abnormal sweating",
      item == 9  ~ "Dry mouth",
      item == 10 ~ "Chest pain"
    ),
    symptom = fct_reorder(symptom, item)
  )
```

## Overview

This document explores how **warned symptoms** (headache and dizziness — explicitly
mentioned in the side effect warning) compare to **unwarned symptoms** (the
remaining 8 GASE items) across experimental conditions. All charts show
**difference scores** (active rating − baseline rating), where positive values
indicate symptom worsening.

::: {.callout-note}
**Study design reminder:** Participants (N = 161) were randomised to five groups:
Natural History control (NH), Instruction alone (NINSM), Instruction + Side-Effect
Modelling (NISM), Instruction + Positive Social Modelling video (INSM), and all
conditions combined (ISM).
:::

## Results {.tabset}

### Overall comparison
```{r tab1}
sig_labels <- df_long %>%
  group_by(symptom, item, warned) %>%
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),
    p_val     = t.test(diff)$p.value,
    .groups   = "drop"
  ) %>%
  mutate(
    sig_label = case_when(
      p_val < .001 ~ "**",
      p_val < .05  ~ "*",
      TRUE         ~ ""
    ),
    label     = paste0(symptom, sig_label),
    bar_color = case_when(
      warned      ~ "#D85A30",
      p_val < .05 ~ "#5DCAA5",
      TRUE        ~ "rgba(136,135,128,0.4)"
    )
  )

plot_ly(
  sig_labels,
  x             = ~fct_reorder(label, item),
  y             = ~mean_diff,
  type          = "bar",
  marker        = list(color = ~bar_color),
  hovertemplate = ~paste0(
    "<b>", symptom, "</b><br>",
    "Mean diff: ", round(mean_diff, 3), "<br>",
    "p = ", round(p_val, 3),
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis         = list(title = "", tickangle = -30),
    yaxis         = list(title = "Mean symptom change (difference score)",
                         zeroline = TRUE, zerolinecolor = "#999"),
    showlegend    = FALSE,
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent"
  )
```

::: {.callout-tip}
**Key finding:** Headache (mean = +0.26, p < .001) and dizziness (mean = +0.61,
p < .001) increased significantly across the full sample. Nearly all unwarned
symptoms hovered near zero or declined — indicating that the nocebo effect is
**highly specific** to symptoms participants were explicitly warned about.

`*` p < .05 &nbsp; `**` p < .001
:::

---

### By experimental group
```{r tab2}
group_composite <- df_long %>%
  group_by(group, warned) %>%
  summarise(mean_diff = mean(diff, na.rm = TRUE), .groups = "drop") %>%
  mutate(group = factor(group, levels = c("NH", "NINSM", "NISM", "INSM", "ISM")))

plot_ly(
  group_composite %>% filter(warned),
  x             = ~group,
  y             = ~mean_diff,
  type          = "bar",
  name          = "Warned",
  marker        = list(color = "#D85A30"),
  hovertemplate = "Warned: %{y:.3f}<extra></extra>"
) %>%
  add_trace(
    data          = group_composite %>% filter(!warned),
    x             = ~group,
    y             = ~mean_diff,
    name          = "Unwarned",
    marker        = list(color = "#5DCAA5"),
    hovertemplate = "Unwarned: %{y:.3f}<extra></extra>"
  ) %>%
  layout(
    barmode       = "group",
    xaxis         = list(title = ""),
    yaxis         = list(title = "Composite mean difference score",
                         zeroline = TRUE, zerolinecolor = "#999"),
    showlegend    = TRUE,
    legend        = list(orientation = "h", x = 0, y = -0.2),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent"
  )
```

::: {.callout-tip}
**Group pattern:** The two no-PSM placebo groups (NINSM, NISM) show the highest
warned symptom scores (+1.18 and +1.56). PSM substantially reduces them. Unwarned
symptoms remain near zero or negative across all groups — the nocebo effect does
**not generalise** to unmentioned symptoms.

Placebo vs NH — warned: Cohen *d* = 0.52, *p* = .017; unwarned: *d* = 0.15, *p* = .419.
:::

---

### PSM intervention effect
```{r tab3}
psm_composite <- df_long %>%
  filter(group != "NH") %>%
  mutate(psm = if_else(intervention_bin == "Int", "PSM", "No PSM")) %>%
  group_by(psm, warned) %>%
  summarise(mean_diff = mean(diff, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    symptom_type = if_else(
      warned,
      "Warned (headache + dizziness)",
      "Unwarned (8-item composite)"
    )
  )

plot_ly(
  psm_composite %>% filter(psm == "No PSM"),
  x             = ~symptom_type,
  y             = ~mean_diff,
  type          = "bar",
  name          = "No PSM",
  marker        = list(color = "rgba(136,135,128,0.45)"),
  hovertemplate = "No PSM: %{y:.3f}<extra></extra>"
) %>%
  add_trace(
    data          = psm_composite %>% filter(psm == "PSM"),
    x             = ~symptom_type,
    y             = ~mean_diff,
    name          = "PSM",
    marker        = list(color = "#1D9E75"),
    hovertemplate = "PSM: %{y:.3f}<extra></extra>"
  ) %>%
  layout(
    barmode       = "group",
    xaxis         = list(title = ""),
    yaxis         = list(title = "Composite mean difference score",
                         zeroline = TRUE, zerolinecolor = "#999"),
    showlegend    = TRUE,
    legend        = list(orientation = "h", x = 0, y = -0.2),
    plot_bgcolor  = "transparent",
    paper_bgcolor = "transparent"
  )
```

::: {.callout-tip}
**Symmetric protection:** PSM reduced warned symptoms by −0.73 and unwarned
symptoms by −0.79 — nearly identical magnitudes. This suggests PSM generates a
**broad psychological protective effect** rather than removing symptom-specific
expectations, possibly via improved overall safety perception of the treatment.
:::