---
title: ""
# subtitle: "One-sentence description of your research question"
# author: "SIDs 560622278"
# date: "S1 2026"
format:
  html:
    title-block-banner: true
    code-fold: true
    embed-resources: true
    toc: true
    toc-location: right
    toc-float: true
    include-before-body: data1901-banner.html
    css: style.css
self-contained: true
execute:
  warning: false
  message: false
editor: visual
---

```{r}
library(tidyverse)
library(readxl)
library(here)

#-----Load data--------------------------------------------------------
datapath <- here("data", "1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

#-----Tidy and compute pre/post change---------------------------------
df_long <- df |>
# filter data
  select(
    gid_pid, group, intervention_bin,
    starts_with("bgase_"), starts_with("agase_")
  ) |>
# change data format from wide to long, so we have:  gid_pid | phase | item | value 
  pivot_longer(
    cols          = starts_with("bgase_") | starts_with("agase_"), # select cols
    names_to      = c("phase", "item"), # separate into two new cols
    names_pattern = "(bgase|agase)_(\\d+)" 
  ) |>
# change to wide, so we have:  gid_pid | item | bgase | agase 
  pivot_wider(names_from = phase, values_from = value) |>
#creating cols
  mutate(
    item = as.integer(item),       # change char "1" to int 1
    diff = agase - bgase,          # calculate difference
    warned = if_else(item <= 2, "Warned", "Non-warned"),          # labeling symptoms
    warned = factor(warned, levels = c("Warned", "Non-warned")),  # order "Warned" before "Non-warned"
    symptom = case_when(           # new col. symptom name 
      item == 1  ~ "Headache",     # if item is "1", assign "Headache"
      item == 2  ~ "Dizziness",
      item == 3  ~ "Nausea",
      item == 4  ~ "Stomach discomfort",
      item == 5  ~ "Breathing problems",
      item == 6  ~ "Palpitation",
      item == 7  ~ "Itching",
      item == 8  ~ "Abnormal sweating",
      item == 9  ~ "Dry mouth",
      item == 10 ~ "Chest pain"
    ),
    # forcats package (from tidyverse) handles factors
    symptom = forcats::fct_reorder(symptom, item), # cast to factor and reorder according to number
    # cast group col to factor and assign order
    group = factor(group, levels = c("NH", "NINSM", "NISM", "INSM", "ISM"))
  )
```

# Executive Summary (Max 50 words)

In this research, we discovered a significant increase in the symptom score for warned symptoms, while a relatively stable score for unwarned symptoms. This supports the nocebo effect, providing evidence for the association between social modeling and the occurrence of side effects.

# Evidence

## EDA (Max 200 words)

The data was obtained in the survey from the article “Positive social modeling attenuates nocebo side effects”. Our analysis examined the quantitative variable of self rated symptom score (1-7), and a combination of qualitative data on symptom types, different groups(ISM, NISM, NINSM, NH, INSM) either warned or non-warned.

### Limitations

The data was collected from mainly young psychological students leading to potential sampling bias and consent bias as these people may be more interested in psychology and therefore more compliant or attentive. Moreover, with the selection bias on removing unwanted volunteers (eg. people with allergy, dietary restrictions and high baseline restriction).

### Leading to issues in

+----------------------------+-----------------------------------------------------------------------------------------------+
| Reliability                | • Self reported symptoms would be inconsistent                                                |
|                            |                                                                                               |
|                            | • Absence of psychological measures can lead to variating results                             |
+----------------------------+-----------------------------------------------------------------------------------------------+
| Accuracy                   | • Self-report bias → participants may over-report symptoms after being warned (nocebo effect) |
|                            |                                                                                               |
|                            | • Selection bias (healthy young students) → not representative of typical patients            |
+----------------------------+-----------------------------------------------------------------------------------------------+
| Interpretation of the data | • Warnings increase reported symptoms                                                         |
+----------------------------+-----------------------------------------------------------------------------------------------+

### Assumptions

We assumed that the participants’ are similar at baseline due to randomisation and their self reported symptom scores are reliable and accurate, the variable of warning causes the increase in symptoms shown than unwarned group excluding impacts of other factors eg. personality, race… if there's an increase. Along with unmentioned symptoms unaffected by expectations.

## Analysis

::: {.callout-note icon="false"}
## Research Question

How do warned symptoms (headache/dizziness) compare to non-warned symptoms?
:::

```{r}
#| label: chart1
#| fig-width: 9
#| fig-height: 5
library(plotly)

# --- Summarise: composite mean diff per group × warned status ----------------
chart1_data <- df_long |>
  group_by(group, warned) |>
  summarise(
    composite_mean_diff = mean(diff, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(warned = recode(warned, "Non-warned" = "Unwarned"))

# --- Plot --------------------------------------------------------------------
p1 <- ggplot(chart1_data,
       aes(x = group, y = composite_mean_diff, fill = warned)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
  scale_fill_manual(
    values = c("Warned" = "#D85A30", "Unwarned" = "#5DCAA5")
  ) +
  labs(
    title = "Composite mean difference score by group",
    y     = "Composite mean difference score",
    x     = NULL,
    fill  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

ggplotly(p1)
```

Figure 1 shows that symptoms are more recognised by the warned group than the unwarned, in general. Specifically noting the NINSM and NISM group having a significant difference in the mean difference score between the warned and unwarned. This suggests that warnings heightened participants’ awareness and sensitivity to possible side effects, especially when paired with social modelling or informational cues. The increase in reported symptoms supports the role of expectancy in amplifying symptom perception (nocebo effect), where negative expectations can increase symptom reporting even without physiological cause (Benedetti, 2013; Colloca & Miller, 2011).

```{r}
#| label: chart2
#| fig-width: 9
#| fig-height: 5
# --- Summarise: one row per symptom with mean change and significance --------
chart2_data <- df_long |>
  group_by(symptom, item, warned) |>       # group by symptom, item number, and warned status
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),  # average pre→post change, ignoring NAs
    p_val     = t.test(diff)$p.value,      # one-sample t-test: does mean diff ≠ 0?
    .groups   = "drop"                     # remove grouping after summarising
  ) |>
  mutate(
    # convert p-value to a star label for the bar chart
    sig_label = case_when(
      p_val < .001 ~ "**",   # highly significant
      p_val < .05  ~ "*",    # significant
      TRUE         ~ ""      # not significant (else case)
    ),
    label = paste0(symptom, sig_label)     # e.g. "Headache**"
  )

# --- Plot --------------------------------------------------------------------
ggplot(chart2_data, aes(x = fct_reorder(label, item), # order bars by item number 1–10
                         y = mean_diff,
                         fill = warned)) +
  geom_col(width = 0.7) +                             # bar chart using actual y values
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) + # reference line at no change
  scale_fill_manual(
    values = c("Warned" = "#D85A30", "Non-warned" = "#5DCAA5"),  # orange-red = warned, teal = non-warned
    labels = c("Warned" = "Warned symptom", "Non-warned" = "Non-warned symptom")
  ) +
  labs(
    title   = "Mean symptom change across all participants",
    x       = NULL,                                    # no x-axis title (labels are self-explanatory)
    y       = "Mean difference score (active − baseline)",
    fill    = NULL,                                    # no legend titlela
    caption = "* p < .05  ** p < .001"                # significance key
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1), # rotate labels to avoid overlap
    legend.position = "top",
    plot.caption    = element_text(colour = "grey50")       # dim caption so it doesn't distract
  )
```

Figure 2 shows that there’s a disparity in the recognition of symptoms between the warned and unwarned group, with the warned group having a significant change in their self scoring of headache and dizziness. With the unwarned group having minimal change and a more wide spread of symptoms being recognised. This indicates that warnings directed participants’ attention toward those specific symptoms, showing selective rather than general increases in symptom recognition. The focused symptom amplification aligns with findings by Faasse and Petrie (2013), who found that side-effect warnings alone can heighten symptom experience through expectation (Faasse & Petrie, 2013).

```{r}
library(tidyverse)
library(here)
library(fmsb)
library(readxl)

datapath <- here("data/1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

before <- c(
  mean(df$bgase_1, na.rm = TRUE),
  mean(df$bgase_2, na.rm = TRUE),
  mean(df$bgase_3, na.rm = TRUE),
  mean(df$bgase_4, na.rm = TRUE),
  mean(df$bgase_5, na.rm = TRUE),
  mean(df$bgase_6, na.rm = TRUE),
  mean(df$bgase_7, na.rm = TRUE),
  mean(df$bgase_8, na.rm = TRUE),
  mean(df$bgase_9, na.rm = TRUE),
  mean(df$bgase_10, na.rm = TRUE)
)

after <- c(
  mean(df$agase_1, na.rm = TRUE),
  mean(df$agase_2, na.rm = TRUE),
  mean(df$agase_3, na.rm = TRUE),
  mean(df$agase_4, na.rm = TRUE),
  mean(df$agase_5, na.rm = TRUE),
  mean(df$agase_6, na.rm = TRUE),
  mean(df$agase_7, na.rm = TRUE),
  mean(df$agase_8, na.rm = TRUE),
  mean(df$agase_9, na.rm = TRUE),
  mean(df$agase_10, na.rm = TRUE)
)

data <- data.frame(
  Headache = c(before[1], after[1]),
  Dizziness = c(before[2], after[2]),
  Nausea = c(before[3], after[3]),
  `Stomach discomfort` = c(before[4], after[4]),
  `Breathing problems` = c(before[5], after[5]),
  Palpitation = c(before[6], after[6]),
  Itching = c(before[7], after[7]),
  `Abnormal sweating` = c(before[8], after[8]),
  `Dry mouth` = c(before[9], after[9]),
  `Chest pain` = c(before[10], after[10])
)

data <- rbind(
  rep(1.8, ncol(data)),
  rep(1, ncol(data)),
  data
)

radarchart(
  data,
  axistype = 1,
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8"),
  pcol = c("steelblue", "tomato"),
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),
  plwd = 1.5,
  plty = c(1, 1),
  cglcol = "grey70",
  axislabcol = "grey30",
  vlcex = 0.8
)

title(
    main = "Mean symptom scores between each symptom: Before vs After",
    col.main = "black",
    cex.main = 1,
    line = 1.5,
    adj  = 0.5   # centre the title horizontally
)

legend(
  "topright",
  legend = c("Before", "After"),
  col = c("steelblue", "tomato"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

Figure 3 indicates the mean symptom scores of the groups before and being warned, having an overall increase in their self scorings. With NINSM and NISM having an increase of 0.4-0.6. Showcasing that exposure to others reporting side effects may have increased participants’ own symptom reporting. This reinforces other research which shows that social modelling intensifies nocebo responses by normalising and validating symptom awareness (Colloca & Miller, 2011).

```{r}
library(tidyverse)
library(here)
library(fmsb)
library(readxl)

datapath <- here("data/1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

before <- c(
    mean(rowMeans(df[df$group == "NH", c("bgase_1", "bgase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "NINSM", c("bgase_1", "bgase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "NISM", c("bgase_1", "bgase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "INSM", c("bgase_1", "bgase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "ISM", c("bgase_1", "bgase_2")],na.rm = TRUE),na.rm = TRUE)
)

after <- c(
  mean(rowMeans(df[df$group == "NH", c("agase_1", "agase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "NINSM", c("agase_1", "agase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "NISM", c("agase_1", "agase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "INSM", c("agase_1", "agase_2")],na.rm = TRUE),na.rm = TRUE),
    mean(rowMeans(df[df$group == "ISM", c("agase_1", "agase_2")],na.rm = TRUE),na.rm = TRUE)
)

data <- data.frame(
  NH = c(before[1], after[1]),
  NINSM = c(before[2], after[2]),
  NISM = c(before[3], after[3]),
  INSM = c(before[4], after[4]),
  ISM = c(before[5], after[5])
)

data <- rbind(
  rep(2, ncol(data)),
  rep(1, ncol(data)),
  data
)

radarchart(
  data,
  seg = 5,
  axistype = 1,
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8", "2.0"),
  pcol = c("steelblue", "tomato"),
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),
  plwd = 1.5,
  plty = c(1, 1),
  cglcol = "grey70",
  axislabcol = "grey30",
  vlcex = 0.8
)

title(
    main = "Mean symptom scores between groups: Before vs After (Warned symptom)",
    col.main = "black",
    cex.main = 1,
    line = 1.5
)

legend(
  "topright",
  legend = c("Before", "After"),
  col = c("steelblue", "tomato"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

Figure 4 shows the mean symptoms of the unwarned group with slight decrease or no change in their symptom scores. The absence of change suggests that physical or environmental factors alone did not influence symptom reporting.

```{r}
library(tidyverse)
library(here)
library(fmsb)
library(readxl)

datapath <- here("data/1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

before <- c(
  mean(rowMeans(df[df$group == "NH",    paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "NINSM", paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "NISM",  paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "INSM",  paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "ISM",   paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE)
)

after <- c(
  mean(rowMeans(df[df$group == "NH",    paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "NINSM", paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "NISM",  paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "INSM",  paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),
  mean(rowMeans(df[df$group == "ISM",   paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE)
)

data <- data.frame(
  NH = c(before[1], after[1]),
  NINSM = c(before[2], after[2]),
  NISM = c(before[3], after[3]),
  INSM = c(before[4], after[4]),
  ISM = c(before[5], after[5])
)

data <- rbind(
  rep(2, ncol(data)),
  rep(1, ncol(data)),
  data
)

radarchart(
  data,
  seg = 5,
  axistype = 1,
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8", "2.0"),
  pcol = c("steelblue", "tomato"),
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),
  plwd = 1.5,
  plty = c(1, 1),
  cglcol = "grey70",
  axislabcol = "grey30",
  vlcex = 0.8
)

title(
    main = "Mean symptom scores between groups: Before vs After (Non-warned symptom)",
    col.main = "black",
    cex.main = 1,
    line = 1.5
)

legend(
  "topright",
  legend = c("Before", "After"),
  col = c("steelblue", "tomato"),
  lty = 1,
  lwd = 2,
  bty = "n"
)
```

Figure 5, showcases the mean symptom scores before and after the experiment, with a sharp rise of dizziness being observed along with a slight increase of headache. Also, having other unwarned symptoms remains constant. Hence, by figure 5 the warned participants have a specific increase in the reporting of those symptoms more than others.

Overall, the consistent increase in warned but not non-warned symptoms demonstrates that expectations strongly influenced symptom experience. Participants were severely influenced by the specific symptoms they anticipated. The findings clearly demonstrate a nocebo effect, however, the study’s limitations may have impacted the results. The sample consisted mainly of psychology students, and data were self-reported, introducing possible bias and limiting generalisability. Despite this, the overall pattern strongly indicates that warning participants about side effects can meaningfully intensify perceived symptom severity.

## Articles (Max 50 words)

### **References**

-   Benedetti, F. (2013) *Placebo and the new physiology of the doctor-patient relationship*. *Physiological Reviews*, 93(3), pp. 1207–1246. doi:10.1152/physrev.00043.2012.

-   Colloca, L. and Miller, F.G. (2011) ‘The nocebo effect and its relevance for clinical practice’, *Psychosomatic Medicine*, 73(7), pp. 598–603. doi:10.1097/PSY.0b013e3182294a50.

-   Faasse, K. and Petrie, K.J. (2013) ‘The nocebo effect: patient expectations and medication side effects’, *Postgraduate Medical Journal*, 89(1055), pp. 540–546. doi:10.1136/postgradmedj-2012-131805.

# Acknowledgements

### Contributions

-   Victoria: Executive summary, research question

-   Joyce: EDA, data analysis

-   Jase: Coding graphs, Quarto report aesthetics

-   Terry: Coding graphs

-   Emily: Research question, connection with other articles

An AI usage statement including: the name and version of the AI tools, the publisher, the URL, a brief description of the context in which the tool was used. If no AI was used, this must be clearly stated.

A short statement explaining how 1 of the Shared Values and 1 of the Ethical Principles has been adhered to.

## AI usage

Claude Opus 4.6, published by Anthropic (<https://claude.ai/>), was used to assist with coding.