---
title: ""
# subtitle: "How do warned symptoms (headache/dizziness) compare to non-warned symptoms?"
# author: "SIDs 560622278"
# date: "S1 2026"
format:
  html:
    title-block-banner: true
    code-fold: true
    embed-resources: true
    toc: true
    toc-location: right
    toc-float: true
    include-before-body: data1901-banner.html
    css: style.css
self-contained: true
execute:
  warning: false
  message: false
editor: visual
---

```{r}
library(tidyverse)
library(readxl)
library(here)
library(fmsb)
library(plotly)


#-----Load data--------------------------------------------------------
datapath <- here("data", "1901_2026_s1_data.xlsx")
df <- read_excel(datapath, sheet = "osf_anon_data")

#-----Tidy and compute pre/post change---------------------------------
df_long <- df |>
# filter data
  select(
    gid_pid, group, intervention_bin,
    starts_with("bgase_"), starts_with("agase_")
  ) |>
# change data format from wide to long, so we have:  gid_pid | phase | item | value 
  pivot_longer(
    cols          = starts_with("bgase_") | starts_with("agase_"), # select cols
    names_to      = c("phase", "item"), # separate into two new cols
    names_pattern = "(bgase|agase)_(\\d+)" 
  ) |>
# change to wide, so we have:  gid_pid | item | bgase | agase 
  pivot_wider(names_from = phase, values_from = value) |>
#creating cols
  mutate(
    item = as.integer(item),       # change char "1" to int 1
    diff = agase - bgase,          # calculate difference
    warned = if_else(item <= 2, "Warned", "Non-warned"),          # labeling symptoms
    warned = factor(warned, levels = c("Warned", "Non-warned")),  # order "Warned" before "Non-warned"
    symptom = case_when(           # new col. symptom name 
      item == 1  ~ "Headache",     # if item is "1", assign "Headache"
      item == 2  ~ "Dizziness",
      item == 3  ~ "Nausea",
      item == 4  ~ "Stomach discomfort",
      item == 5  ~ "Breathing problems",
      item == 6  ~ "Palpitation",
      item == 7  ~ "Itching",
      item == 8  ~ "Abnormal sweating",
      item == 9  ~ "Dry mouth",
      item == 10 ~ "Chest pain"
    ),
    # forcats package (from tidyverse) handles factors
    symptom = forcats::fct_reorder(symptom, item), # cast to factor and reorder according to number
    # cast group col to factor and assign order
    group = factor(group, levels = c("NH", "NINSM", "NISM", "INSM", "ISM"))
  )
```

# Executive Summary

In this research, we discovered a significant increase in the symptom score for warned symptoms, while a relatively stable score for unwarned symptoms. This supports the nocebo effect, providing evidence for the association between social modeling and the occurrence of side effects.

# Evidence

## Exploratory Data Analysis (EDA)

The data was obtained in the survey from the article “Positive social modeling attenuates nocebo side effects”. Our analysis examined the quantitative variable of self rated symptom score (1-7), and a combination of qualitative data on symptom types, different groups(NH, ISM, NISM, INSM, NINSM) either warned or non-warned.

*NH: Natural History\
ISM: Instruction & Side-effect Modelling\
NISM: No Instruction & Side-effect Modelling\
INSM: Instruction & No Side-effect Modelling\
NINSM: No Instruction & No Side-effect Modelling*

### Limitations

The data was collected from mainly young psychological students leading to potential sampling bias and consent bias as these people may be more interested in psychology and therefore more compliant or attentive. Moreover, with the selection bias on removing unwanted volunteers (eg. people with allergy, dietary restrictions and high baseline restriction).

### Leading to issues in

+----------------------------+-----------------------------------------------------------------------------------------------+
| Reliability                | • Self reported symptoms would be inconsistent                                                |
|                            |                                                                                               |
|                            | • Absence of psychological measures can lead to variating results                             |
+----------------------------+-----------------------------------------------------------------------------------------------+
| Accuracy                   | • Self-report bias → participants may over-report symptoms after being warned (nocebo effect) |
|                            |                                                                                               |
|                            | • Selection bias (healthy young students) → not representative of typical patients            |
+----------------------------+-----------------------------------------------------------------------------------------------+
| Interpretation of the data | • Warnings increase reported symptoms                                                         |
+----------------------------+-----------------------------------------------------------------------------------------------+

### Assumptions

We assumed that the participants’ are similar at baseline due to randomisation and their self reported symptom scores are reliable and accurate, the variable of warning causes the increase in symptoms shown than unwarned group excluding impacts of other factors eg. personality, race… if there's an increase. Along with unmentioned symptoms unaffected by expectations.

## Analysis

::: {.callout-note icon="false"}
## Research Question

How do warned symptoms (headache/dizziness) compare to non-warned symptoms?
:::

```{r}
#| label: chart2
#| fig-width: 9
#| fig-height: 6
# --- Summarise: one row per symptom with mean change and significance --------
chart2_data <- df_long |>
  group_by(symptom, item, warned) |>       # group by symptom, item number, and warned status
  summarise(
    mean_diff = mean(diff, na.rm = TRUE),  # average pre→post change, ignoring NAs
    p_val     = t.test(diff)$p.value,      # one-sample t-test: does mean diff ≠ 0?
    .groups   = "drop"                     # remove grouping after summarising
  ) |>
  mutate(
    # convert p-value to a star label for the bar chart
    sig_label = case_when(
      p_val < .001 ~ "**",   # highly significant
      p_val < .05  ~ "*",    # significant
      TRUE         ~ ""      # not significant (else case)
    ),
    label = paste0(symptom, sig_label)     # e.g. "Headache**"
  )

# --- Plot --------------------------------------------------------------------
p2 <- ggplot(chart2_data, aes(x = fct_reorder(label, item), # order bars by item number 1–10
                         y = mean_diff,
                         fill = warned)) +
  geom_col(width = 0.7) +                             # bar chart using actual y values
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) + # reference line at no change
  scale_fill_manual(
    values = c("Warned" = "#D85A30", "Non-warned" = "#5DCAA5"),  # orange-red = warned, teal = non-warned
    labels = c("Warned" = "Warned symptom", "Non-warned" = "Non-warned symptom")
  ) +
  labs(
    x       = NULL,                                    # no x-axis title (labels are self-explanatory)
    y       = "Mean difference score (active − baseline)",
    fill    = NULL,                                    # no legend titlela
    caption = "* p < .05  ** p < .001"                # significance key
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 35, hjust = 1), # rotate labels to avoid overlap
    legend.position = "top",
    plot.caption    = element_text(colour = "grey50")       # dim caption so it doesn't distract
  )

ggplotly(p2)
```

Figure 1 shows that there’s a disparity in the recognition of symptoms between the warned and unwarned group, with the warned group having a significant change in their self scoring of headache and dizziness. With the unwarned group having minimal change and a more wide spread of symptoms being recognised. This indicates that warnings directed participants’ attention toward those specific symptoms, showing selective rather than general increases in symptom recognition. The focused symptom amplification aligns with findings by Faasse and Petrie (2013), who found that side-effect warnings alone can heighten symptom experience through expectation (Faasse & Petrie, 2013).

```{r}
datapath <- here("data/1901_2026_s1_data.xlsx")   # build the file path to the Excel file
df <- read_excel(datapath, sheet = "osf_anon_data")   # read the chosen sheet into a data frame

before <- c(   # create a vector of mean "before" scores for the 10 symptoms
  mean(df$bgase_1, na.rm = TRUE),   # mean of symptom 1 before, ignore missing values
  mean(df$bgase_2, na.rm = TRUE),   # mean of symptom 2 before
  mean(df$bgase_3, na.rm = TRUE),   # mean of symptom 3 before
  mean(df$bgase_4, na.rm = TRUE),   # mean of symptom 4 before
  mean(df$bgase_5, na.rm = TRUE),   # mean of symptom 5 before
  mean(df$bgase_6, na.rm = TRUE),   # mean of symptom 6 before
  mean(df$bgase_7, na.rm = TRUE),   # mean of symptom 7 before
  mean(df$bgase_8, na.rm = TRUE),   # mean of symptom 8 before
  mean(df$bgase_9, na.rm = TRUE),   # mean of symptom 9 before
  mean(df$bgase_10, na.rm = TRUE)   # mean of symptom 10 before
)

after <- c(   # create a vector of mean "after" scores for the 10 symptoms
  mean(df$agase_1, na.rm = TRUE),   # mean of symptom 1 after, ignore missing values
  mean(df$agase_2, na.rm = TRUE),   # mean of symptom 2 after
  mean(df$agase_3, na.rm = TRUE),   # mean of symptom 3 after
  mean(df$agase_4, na.rm = TRUE),   # mean of symptom 4 after
  mean(df$agase_5, na.rm = TRUE),   # mean of symptom 5 after
  mean(df$agase_6, na.rm = TRUE),   # mean of symptom 6 after
  mean(df$agase_7, na.rm = TRUE),   # mean of symptom 7 after
  mean(df$agase_8, na.rm = TRUE),   # mean of symptom 8 after
  mean(df$agase_9, na.rm = TRUE),   # mean of symptom 9 after
  mean(df$agase_10, na.rm = TRUE)   # mean of symptom 10 after
)

data <- data.frame(   # make a data frame with one column per symptom
  Headache = c(before[1], after[1]),   # combine before and after mean for Headache
  Dizziness = c(before[2], after[2]),   # combine before and after mean for Dizziness
  Nausea = c(before[3], after[3]),   # combine before and after mean for Nausea
  `Stomach discomfort` = c(before[4], after[4]),   # combine before and after mean for Stomach discomfort
  `Breathing problems` = c(before[5], after[5]),   # combine before and after mean for Breathing problems
  Palpitation = c(before[6], after[6]),   # combine before and after mean for Palpitation
  Itching = c(before[7], after[7]),   # combine before and after mean for Itching
  `Abnormal sweating` = c(before[8], after[8]),   # combine before and after mean for Abnormal sweating
  `Dry mouth` = c(before[9], after[9]),   # combine before and after mean for Dry mouth
  `Chest pain` = c(before[10], after[10])   # combine before and after mean for Chest pain
)

data <- rbind(   # add two rows at the top for radar chart max and min scale
  rep(1.8, ncol(data)),   # top row sets the maximum value for each axis
  rep(1, ncol(data)),   # second row sets the minimum value for each axis
  data   # actual before and after values
)

radarchart(   # draw the radar chart
  data,
  axistype = 1,   # show axis labels in the default style
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8"),   # labels for the circular grid levels
  pcol = c("steelblue", "tomato"),   # line colours for before and after
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),   # fill colours with transparency
  plwd = 1.5,   # line width
  plty = c(1, 1),   # line type (solid lines)
  cglcol = "grey70",   # colour of the grid lines
  axislabcol = "grey30",   # colour of the axis number labels
  vlcex = 0.8   # size of the symptom labels around the chart
)

title(
  main = "Mean symptom scores between each symptom: Before vs After",   # add main title
  col.main = "black",   # title colour
  cex.main = 1,   # title size
  line = 1.5,   # move title a bit away from the plot
  adj  = 0.5   # centre the title horizontally
)

legend(
  "topright",   # place legend at the top right
  legend = c("Before", "After"),   # legend labels
  col = c("steelblue", "tomato"),   # legend line colours
  lty = 1,   # solid line type in legend
  lwd = 2,   # line width in legend
  bty = "n"   # remove legend box border
)
```

Figure 2 indicates the mean symptom scores of the groups before and being warned, having an overall increase in their self scorings. With NINSM and NISM having an increase of 0.4-0.6. Showcasing that exposure to others reporting side effects may have increased participants’ own symptom reporting. This reinforces other research which shows that social modelling intensifies nocebo responses by normalising and validating symptom awareness (Colloca & Miller, 2011).

```{r}
#| label: chart1
#| fig-width: 9
#| fig-height: 6

# --- Summarise: composite mean diff per group × warned status ----------------
chart1_data <- df_long |>
  group_by(group, warned) |>
  summarise(
    composite_mean_diff = mean(diff, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(warned = recode(warned, "Non-warned" = "Unwarned"))

# --- Plot --------------------------------------------------------------------
p1 <- ggplot(chart1_data,
       aes(x = group, y = composite_mean_diff, fill = warned)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
  scale_fill_manual(
    values = c("Warned" = "#D85A30", "Unwarned" = "#5DCAA5")
  ) +
  labs(
    y     = "Composite mean difference score",
    x     = NULL,
    fill  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

ggplotly(p1)
```

Figure 3 shows that symptoms are more recognised by the warned group than the unwarned, in general. Specifically noting the NINSM and NISM group having a significant difference in the mean difference score between the warned and unwarned. This suggests that warnings heightened participants’ awareness and sensitivity to possible side effects, especially when paired with social modelling or informational cues. The increase in reported symptoms supports the role of expectancy in amplifying symptom perception (nocebo effect), where negative expectations can increase symptom reporting even without physiological cause (Benedetti, 2013; Colloca & Miller, 2011).

```{r}
datapath <- here("data/1901_2026_s1_data.xlsx")   # build the file path to the Excel file
df <- read_excel(datapath, sheet = "osf_anon_data")   # read the chosen sheet into a data frame

before <- c(   # create a vector of mean "before" warned-symptom scores for each group
  mean(rowMeans(df[df$group == "NH", c("bgase_1", "bgase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness before for NH
  mean(rowMeans(df[df$group == "NINSM", c("bgase_1", "bgase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness before for NINSM
  mean(rowMeans(df[df$group == "NISM", c("bgase_1", "bgase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness before for NISM
  mean(rowMeans(df[df$group == "INSM", c("bgase_1", "bgase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness before for INSM
  mean(rowMeans(df[df$group == "ISM", c("bgase_1", "bgase_2")], na.rm = TRUE), na.rm = TRUE)   # mean of Headache and Dizziness before for ISM
)

after <- c(   # create a vector of mean "after" warned-symptom scores for each group
  mean(rowMeans(df[df$group == "NH", c("agase_1", "agase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness after for NH
  mean(rowMeans(df[df$group == "NINSM", c("agase_1", "agase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness after for NINSM
  mean(rowMeans(df[df$group == "NISM", c("agase_1", "agase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness after for NISM
  mean(rowMeans(df[df$group == "INSM", c("agase_1", "agase_2")], na.rm = TRUE), na.rm = TRUE),   # mean of Headache and Dizziness after for INSM
  mean(rowMeans(df[df$group == "ISM", c("agase_1", "agase_2")], na.rm = TRUE), na.rm = TRUE)   # mean of Headache and Dizziness after for ISM
)

data <- data.frame(   # make a data frame with one column per group
  NH = c(before[1], after[1]),   # combine before and after mean for NH
  NINSM = c(before[2], after[2]),   # combine before and after mean for NINSM
  NISM = c(before[3], after[3]),   # combine before and after mean for NISM
  INSM = c(before[4], after[4]),   # combine before and after mean for INSM
  ISM = c(before[5], after[5])   # combine before and after mean for ISM
)

data <- rbind(   # add two rows at the top for radar chart max and min scale
  rep(2, ncol(data)),   # top row sets the maximum value for each axis
  rep(1, ncol(data)),   # second row sets the minimum value for each axis
  data   # actual before and after values
)

radarchart(
  data,   # radar chart input data
  seg = 5,   # split the axis into 5 segments
  axistype = 1,   # show axis labels in the default style
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8", "2.0"),   # labels for the circular grid levels
  pcol = c("steelblue", "tomato"),   # line colours for before and after
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),   # fill colours with transparency
  plwd = 1.5,   # line width
  plty = c(1, 1),   # line type (solid lines)
  cglcol = "grey70",   # colour of the grid lines
  axislabcol = "grey30",   # colour of the axis number labels
  vlcex = 0.8   # size of the group labels around the chart
)

title(
  main = "Mean symptom scores across groups: Before vs After (Warned symptoms)",   # add main title
  col.main = "black",   # title colour
  cex.main = 1,   # title size
  line = 1.5   # move title slightly away from the plot
)

legend(
  "topright",   # place legend at the top right
  legend = c("Before", "After"),   # legend labels
  col = c("steelblue", "tomato"),   # legend line colours
  lty = 1,   # solid line type in legend
  lwd = 2,   # line width in legend
  bty = "n"   # remove legend box border
)
```

Figure 4 shows the mean symptoms of the unwarned group with slight decrease or no change in their symptom scores. The absence of change suggests that physical or environmental factors alone did not influence symptom reporting.

```{r}
datapath <- here("data/1901_2026_s1_data.xlsx")   # build the file path to the Excel file
df <- read_excel(datapath, sheet = "osf_anon_data")   # read the chosen sheet into a data frame

before <- c(   # create a vector of mean "before" non-warned-symptom scores for each group
  mean(rowMeans(df[df$group == "NH",    paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 before for NH
  mean(rowMeans(df[df$group == "NINSM", paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 before for NINSM
  mean(rowMeans(df[df$group == "NISM",  paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 before for NISM
  mean(rowMeans(df[df$group == "INSM",  paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 before for INSM
  mean(rowMeans(df[df$group == "ISM",   paste0("bgase_", 3:10)], na.rm = TRUE), na.rm = TRUE)    # mean of symptoms 3 to 10 before for ISM
)

after <- c(   # create a vector of mean "after" non-warned-symptom scores for each group
  mean(rowMeans(df[df$group == "NH",    paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 after for NH
  mean(rowMeans(df[df$group == "NINSM", paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 after for NINSM
  mean(rowMeans(df[df$group == "NISM",  paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 after for NISM
  mean(rowMeans(df[df$group == "INSM",  paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE),   # mean of symptoms 3 to 10 after for INSM
  mean(rowMeans(df[df$group == "ISM",   paste0("agase_", 3:10)], na.rm = TRUE), na.rm = TRUE)    # mean of symptoms 3 to 10 after for ISM
)

data <- data.frame(   # make a data frame with one column per group
  NH = c(before[1], after[1]),         # combine before and after mean for NH
  NINSM = c(before[2], after[2]),      # combine before and after mean for NINSM
  NISM = c(before[3], after[3]),       # combine before and after mean for NISM
  INSM = c(before[4], after[4]),       # combine before and after mean for INSM
  ISM = c(before[5], after[5])         # combine before and after mean for ISM
)

data <- rbind(   # add two rows at the top for radar chart max and min scale
  rep(2, ncol(data)),   # top row sets the maximum value for each axis
  rep(1, ncol(data)),   # second row sets the minimum value for each axis
  data                  # actual before and after values
)

radarchart(
  data,   # radar chart input data
  seg = 5,   # split each axis into 5 segments
  axistype = 1,   # show axis labels in the default style
  caxislabels = c("1.0", "1.2", "1.4", "1.6", "1.8", "2.0"),   # labels for the circular grid levels
  pcol = c("steelblue", "tomato"),   # line colours for before and after
  pfcol = c(scales::alpha("steelblue", 0.2), scales::alpha("tomato", 0.2)),   # fill colours with transparency
  plwd = 1.5,   # line width
  plty = c(1, 1),   # line type (solid lines)
  cglcol = "grey70",   # colour of the grid lines
  axislabcol = "grey30",   # colour of the axis number labels
  vlcex = 0.8   # size of the group labels around the chart
)

title(
  main = "Mean symptom scores across groups: Before vs After (Non-warned symptoms)",   # add main title
  col.main = "black",   # title colour
  cex.main = 1,   # title size
  line = 1.5   # move title slightly away from the plot
)

legend(
  "topright",   # place legend at the top right
  legend = c("Before", "After"),   # legend labels
  col = c("steelblue", "tomato"),   # legend line colours
  lty = 1,   # solid line type in legend
  lwd = 2,   # line width in legend
  bty = "n"   # remove legend box border
)
```

Figure 5, showcases the mean symptom scores before and after the experiment, with a sharp rise of dizziness being observed along with a slight increase of headache. Also, having other unwarned symptoms remains constant. Hence, by figure 5 the warned participants have a specific increase in the reporting of those symptoms more than others.

Overall, the consistent increase in warned but not non-warned symptoms demonstrates that expectations strongly influenced symptom experience. Participants were severely influenced by the specific symptoms they anticipated. The findings clearly demonstrate a nocebo effect, however, the study’s limitations may have impacted the results. The sample consisted mainly of psychology students, and data were self-reported, introducing possible bias and limiting generalisability. Despite this, the overall pattern strongly indicates that warning participants about side effects can meaningfully intensify perceived symptom severity.

## Articles

### **References**

-   Benedetti, F. (2013) *Placebo and the new physiology of the doctor-patient relationship*. *Physiological Reviews*, 93(3), pp. 1207–1246. doi:10.1152/physrev.00043.2012.

-   Colloca, L. and Miller, F.G. (2011) ‘The nocebo effect and its relevance for clinical practice’, *Psychosomatic Medicine*, 73(7), pp. 598–603. doi:10.1097/PSY.0b013e3182294a50.

-   Faasse, K. and Petrie, K.J. (2013) ‘The nocebo effect: patient expectations and medication side effects’, *Postgraduate Medical Journal*, 89(1055), pp. 540–546. doi:10.1136/postgradmedj-2012-131805.

# Acknowledgements

## Tools used

+-------------------------------------------+-----------------------+
| **Communication**                         | Instagram             |
+-------------------------------------------+-----------------------+
| **Writing**                               | Notion, Google Docs   |
+-------------------------------------------+-----------------------+
| **Editing**                               | Positron              |
+-------------------------------------------+-----------------------+
| **File management & version** **control** | OneDrive, GitHub      |
+-------------------------------------------+-----------------------+

## Contributions

+--------------+------------------------------------------------------------------------+
| **Victoria** | Executive summary, research question                                   |
+--------------+------------------------------------------------------------------------+
| **Joyce**    | EDA, data analysis                                                     |
+--------------+------------------------------------------------------------------------+
| **Jase**     | Coding graphs, Quarto report aesthetics and polishing                  |
+--------------+------------------------------------------------------------------------+
| **Terry**    | Coding graphs, presentation slides polishing                           |
+--------------+------------------------------------------------------------------------+
| **Emily**    | Research question, connection with other articles, presentation slides |
+--------------+------------------------------------------------------------------------+

[***A short statement explaining how 1 of the Shared Values and 1 of the Ethical Principles has been adhered to.***]{.underline}

## AI Usage

Claude Opus 4.6, published by Anthropic (<https://claude.ai/>), was used to assist with coding.

ChatGPT-5.3, published by Open AI (<https://chatgpt.com/),> was used to assist with coding, understanding bias and reliability relationships.