---
title: "DATA1901 Project 1: Student Lifestyle and Academic Performance"
author:
  - name: "Group Member 1"
  - name: "Group Member 2"
  - name: "Group Member 3"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-tools: true
  pdf:
    documentclass: article
    toc: true
    number-sections: true
    colorlinks: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(corrplot)
```

# Introduction

This report investigates the relationship between university students' lifestyle
habits and their academic performance (GPA). Using survey data collected from
100 students across multiple faculties, we explore how factors such as weekly
study hours, sleep, exercise, part-time employment, and commute time influence
academic outcomes.

**Research Questions:**

1. Is there a significant association between weekly study hours and GPA?
2. How do sleep and exercise habits relate to academic performance?
3. Does part-time employment affect GPA and study time?
4. Are there differences in academic performance across faculties?

# Data Description

The dataset was collected via an anonymous online survey distributed to
undergraduate students. It contains 100 observations and 12 variables.

```{r}
#| label: load-data

# Load the dataset
students <- read_csv("data/student_survey.csv", show_col_types = FALSE)

# Convert categorical variables to factors
students <- students |>
  mutate(
    gender         = factor(gender),
    part_time_job  = factor(part_time_job, levels = c("No", "Yes")),
    faculty        = factor(faculty),
    year_of_study  = factor(year_of_study, levels = 1:4,
                            labels = c("Year 1", "Year 2",
                                       "Year 3", "Year 4"))
  )
```

## Dataset Overview

```{r}
#| label: tbl-overview
#| tbl-cap: "First six rows of the student survey dataset"

students |>
  head(6) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for numeric variables"

students |>
  select(age, study_hours_per_week, sleep_hours_per_night,
         exercise_days_per_week, commute_minutes, gpa,
         satisfaction_score) |>
  summary() |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Dataset dimensions:** `r nrow(students)` observations across
`r ncol(students)` variables.

Variable descriptions:

| Variable | Type | Description |
|---|---|---|
| `student_id` | Integer | Unique student identifier |
| `age` | Integer | Student age (years) |
| `gender` | Categorical | Gender (Female/Male) |
| `study_hours_per_week` | Integer | Weekly study hours outside class |
| `sleep_hours_per_night` | Numeric | Average nightly sleep (hours) |
| `exercise_days_per_week` | Integer | Exercise days per week |
| `part_time_job` | Categorical | Has a part-time job (Yes/No) |
| `commute_minutes` | Integer | One-way commute time (minutes) |
| `gpa` | Numeric | Grade Point Average (0–7 scale) |
| `faculty` | Categorical | Faculty of enrolment |
| `year_of_study` | Integer | Year of study (1–4) |
| `satisfaction_score` | Integer | Life satisfaction (1–10 scale) |

: Variable descriptions {tbl-colwidths="[25,15,60]"}

# Exploratory Data Analysis

## Distribution of GPA

```{r}
#| label: fig-gpa-dist
#| fig-cap: "Distribution of GPA among surveyed students"
#| fig-alt: "Histogram showing a roughly normal distribution of GPA values
#|   centred around 6.0"

ggplot(students, aes(x = gpa)) +
  geom_histogram(binwidth = 0.2, fill = "#2C7BB6", colour = "white",
                 alpha = 0.85) +
  geom_vline(aes(xintercept = mean(gpa)), colour = "#D7191C",
             linetype = "dashed", linewidth = 1) +
  labs(
    x     = "GPA (0–7 scale)",
    y     = "Number of Students",
    title = "Distribution of Student GPA",
    caption = paste0("Red dashed line = mean GPA (",
                     round(mean(students$gpa), 2), ")")
  ) +
  theme_minimal(base_size = 13)
```

The mean GPA in the sample is **`r round(mean(students$gpa), 2)`**
(SD = `r round(sd(students$gpa), 2)`), with values ranging from
`r min(students$gpa)` to `r max(students$gpa)`.

## GPA by Faculty

```{r}
#| label: fig-gpa-faculty
#| fig-cap: "Boxplot of GPA by faculty"
#| fig-alt: "Side-by-side boxplots comparing GPA distributions across five
#|   faculties"

students |>
  mutate(faculty = fct_reorder(faculty, gpa, .fun = median)) |>
  ggplot(aes(x = faculty, y = gpa, fill = faculty)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x     = "Faculty",
    y     = "GPA",
    title = "GPA Distribution by Faculty"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r}
#| label: tbl-gpa-faculty
#| tbl-cap: "Mean and median GPA by faculty"

students |>
  group_by(faculty) |>
  summarise(
    n          = n(),
    Mean_GPA   = round(mean(gpa), 2),
    Median_GPA = round(median(gpa), 2),
    SD_GPA     = round(sd(gpa), 2)
  ) |>
  arrange(desc(Mean_GPA)) |>
  kable(col.names = c("Faculty", "n", "Mean GPA",
                      "Median GPA", "SD")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## Study Hours and GPA

```{r}
#| label: fig-study-gpa
#| fig-cap: "Scatter plot of weekly study hours versus GPA, coloured by
#|   part-time job status"
#| fig-alt: "Scatter plot showing a positive linear relationship between
#|   study hours and GPA"

ggplot(students, aes(x = study_hours_per_week, y = gpa,
                     colour = part_time_job)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_colour_manual(values = c("No" = "#2C7BB6", "Yes" = "#D7191C"),
                      name   = "Part-time job") +
  labs(
    x     = "Study Hours per Week",
    y     = "GPA",
    title = "Study Hours vs GPA by Part-time Employment Status"
  ) +
  theme_minimal(base_size = 13)
```

## Sleep, Exercise, and Academic Performance

```{r}
#| label: fig-sleep-exercise
#| fig-cap: "Scatter plot of sleep hours vs GPA, faceted by exercise level"
#| fig-alt: "Faceted scatter plots showing GPA trends across different
#|   exercise levels"

students |>
  mutate(exercise_group = cut(exercise_days_per_week,
                              breaks  = c(-Inf, 1, 3, Inf),
                              labels  = c("Low (0–1 days)",
                                          "Moderate (2–3 days)",
                                          "High (4+ days)"))) |>
  ggplot(aes(x = sleep_hours_per_night, y = gpa, colour = exercise_group)) +
  geom_point(alpha = 0.65, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~exercise_group) +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    x     = "Sleep Hours per Night",
    y     = "GPA",
    title = "Sleep vs GPA by Exercise Level"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

# Statistical Analysis

## Research Question 1: Study Hours and GPA

We fit a simple linear regression to quantify the association between weekly
study hours and GPA.

```{r}
#| label: tbl-lm-study
#| tbl-cap: "Simple linear regression: GPA ~ study hours per week"

model_study <- lm(gpa ~ study_hours_per_week, data = students)
model_study |>
  summary() |>
  coef() |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(col.names = c("Term", "Estimate", "Std. Error",
                      "t value", "Pr(>|t|)")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

The slope estimate indicates that each additional study hour per week is
associated with a **`r round(coef(model_study)[2], 3)`** unit increase in GPA
(*p* < 0.001). The model explains
**`r round(summary(model_study)$r.squared * 100, 1)`%** of the variance in GPA
(R² = `r round(summary(model_study)$r.squared, 3)`).

## Research Question 2: Part-time Employment and GPA

```{r}
#| label: fig-ptjob-gpa
#| fig-cap: "GPA by part-time job status with individual data points"

ggplot(students, aes(x = part_time_job, y = gpa, fill = part_time_job)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(width = 0.07, alpha = 0.4, size = 1.5) +
  scale_fill_manual(values = c("No" = "#2C7BB6", "Yes" = "#D7191C")) +
  labs(
    x     = "Has Part-time Job",
    y     = "GPA",
    title = "GPA by Part-time Employment Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

```{r}
#| label: ptjob-ttest

t_test_result <- t.test(gpa ~ part_time_job, data = students)
t_test_result
```

A Welch two-sample *t*-test shows that students without a part-time job have a
significantly higher mean GPA
(*M* = `r round(t_test_result$estimate[1], 2)`) compared to those with a
part-time job
(*M* = `r round(t_test_result$estimate[2], 2)`),
*t*(`r round(t_test_result$parameter, 1)`) =
`r round(t_test_result$statistic, 2)`,
*p* = `r round(t_test_result$p.value, 4)`.

## Research Question 3: Correlation Analysis

```{r}
#| label: fig-corr
#| fig-cap: "Correlation matrix of numeric variables"
#| fig-alt: "Correlation matrix heatmap showing pairwise correlations
#|   between all numeric variables"

numeric_vars <- students |>
  select(age, study_hours_per_week, sleep_hours_per_night,
         exercise_days_per_week, commute_minutes, gpa,
         satisfaction_score)

cor_matrix <- cor(numeric_vars, use = "complete.obs")

corrplot(cor_matrix,
         method  = "color",
         type    = "upper",
         addCoef.col = "black",
         tl.col  = "black",
         tl.srt  = 45,
         number.cex = 0.75,
         col     = colorRampPalette(c("#D7191C", "white", "#2C7BB6"))(200),
         title   = "Pairwise Correlations",
         mar     = c(0, 0, 1, 0))
```

```{r}
#| label: tbl-corr-gpa
#| tbl-cap: "Pearson correlations with GPA"

cor_with_gpa <- cor_matrix[, "gpa"]
cor_with_gpa <- cor_with_gpa[names(cor_with_gpa) != "gpa"]

data.frame(
  Variable    = names(cor_with_gpa),
  Correlation = round(cor_with_gpa, 3)
) |>
  arrange(desc(abs(Correlation))) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## Multiple Linear Regression

To account for multiple predictors simultaneously, we fit a multiple linear
regression model with GPA as the outcome.

```{r}
#| label: tbl-mlr
#| tbl-cap: "Multiple linear regression: GPA ~ lifestyle variables"

model_mlr <- lm(
  gpa ~ study_hours_per_week + sleep_hours_per_night +
    exercise_days_per_week + part_time_job + commute_minutes,
  data = students
)

model_mlr |>
  summary() |>
  coef() |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(col.names = c("Term", "Estimate", "Std. Error",
                      "t value", "Pr(>|t|)")) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

The multiple regression model explains
**`r round(summary(model_mlr)$r.squared * 100, 1)`%** of the variance in GPA
(adjusted R² = `r round(summary(model_mlr)$adj.r.squared, 3)`).

# Discussion

The analysis reveals several key findings:

1. **Study hours** are the strongest predictor of GPA, with each additional
   hour of weekly study associated with a ~`r round(coef(model_study)[2], 2)`
   GPA increase. This aligns with the widely documented relationship between
   time-on-task and academic achievement.

2. **Part-time employment** is associated with lower GPA. Students holding
   part-time jobs tend to study fewer hours per week, which may partly explain
   this relationship (i.e., time displacement).

3. **Sleep and exercise** show moderate positive correlations with GPA,
   suggesting that lifestyle behaviours beyond study time also matter for
   academic outcomes.

4. **Commute time** shows a modest negative correlation with GPA, consistent
   with the idea that longer commutes reduce the time available for study and
   rest.

**Limitations:** The data are self-reported and cross-sectional, so causal
inference is not warranted. The sample (n = 100) may not be representative of
the broader student population, and confounding variables (e.g., prior academic
ability, course difficulty) are not captured.

# Conclusion

This study examined the association between lifestyle factors and academic
performance in university students. Study hours emerge as the most important
predictor of GPA. Part-time employment, sleep, and exercise also relate to
academic outcomes in meaningful ways. Future work could employ a longitudinal
design or larger sample to strengthen causal interpretation.

# References

- R Core Team (2024). *R: A Language and Environment for Statistical
  Computing*. R Foundation for Statistical Computing, Vienna, Austria.
  <https://www.R-project.org/>
- Allaire, J. J. et al. (2024). *Quarto* (Version 1.4).
  <https://quarto.org>
- Wickham, H. et al. (2019). Welcome to the tidyverse. *Journal of Open
  Source Software*, 4(43), 1686. <https://doi.org/10.21105/joss.01686>
- Wei, T. & Simko, V. (2021). *corrplot: Visualization of a Correlation
  Matrix*. R package version 0.92.
  <https://github.com/taiyun/corrplot>
